---
title: "Cyclomort"
author: "Elie Gurarie, Peter Thompson"
date: "2019-02-22"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
require(cyclomort)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

##Cyclomort: an R package to fit periodic survival curves to right-censored mortality data

Survival modelling is one of the most important subfields of ecology. Understanding the mortality rates of a specific population of animals, and how those mortality rates may change over time, can be incredibly important to preserving the survival of these species. In many cases, the risk of mortality at any given time for an individual (the hazard function) varies markedly over the course of a year. For example, migratory species are usually much more susceptible to death during the period when they are migrating. Humans are also more likely to be infected with a variety of bacterial and viral pathogens during the colder months. Being able to model these trends and identify peaks and periodicities in mortality using known mortality data will allow scientists to better solve ecological and medical problems facing the world today.
  
This package is the first R package that uses maximum likelihood theory to produce parametric estimates for a hazard function with any number of periodicities in a single year. We include functions that return parametric fits given a set of mortality data (times of death or censoring); in order to confirm that our fitting procedure is accurate, we also include the option to simulate periodic mortality data with a given mortality function and analyze the MLEs for each parameter in the mortality function. The functions from this package can be used in conjunction with the widely used "survival" package, and the fit_cyclomort function takes these "Surv" objects as input parameters. While small, this package provides a useful framework for estimating the parameters of a periodic mortality function, and has a wide variety of applications in survival modelling.

##Producing a probability function to represent a periodic mortality curve

To generate a function with multiple, differently shaped peaks in one "period" (allowing for the highest level of customizability for users), we adapted a simple version of the wrapped Cauchy distribution that allows for multiple peaks. This "mixture model" incorporates any number of individual Cauchy distributions, each with an individual mean parameter (the location of the peak of the distribution) as well as a concentration parameter (referred to here as rho; this value ranges from 0 to 1, with a higher rho implying that more data is concentrated around the mean). Because we are mixing multiple individual distributions, we must also specify the corresponding weights for each component of the mixuture. This function is wrapped around a period; i.e., it has a bounded distribution on every period with the same shape.

```{r, fig.width = 7, fig.align = "center", echo = FALSE}
plot(seq(-6, 6, .01/pi), DwrappedMultiCauchy(seq(-pi*6, pi * 6, .01), A = 1, mus = c(-1,1), rhos = c(0.7,0.5), weights = 0.7), type = "l", xlab = "x-value (in terms of pi)", ylab = "Probability", main = "peaks = c(-1,1), rhos = c(0.7,0.5), weights = c(0.7,0.3)")
```

```{r, fig.width = 7, fig.align = "center", echo = FALSE}
plot(seq(-6, 6, .01/pi), DwrappedMultiCauchy(seq(-pi*6, pi * 6, .01), A = 1, mus = c(-1, 1, 3), rhos = c(0.4, 0.9, 0.6), weights = c(0.4,0.3)), type = "l", xlab = "x-value (in terms of pi)", ylab = "Probability", main = "peaks = c(-1,1,3), rhos = c(0.4,0.9,0.6), weights = c(0.4,0.3,0.3)")
```

Here the effects of the parameters on the graph become quite clear. Note how the components of the mixture with higher rho values have distributions with more data closer to the peak. Note how in the second graph, even though the weights are distributed relatively equally, the second peak is much higher (albeit there is a shorter "window" of high mortality) due to the very high rho value of 0.9.

##The fit_cyclomort function: parametric estimates for periodic mortality functions

Given a set of mortality data (formatted as a list of times of death/censoring with information about the current state (dead/censored) of each observation), the fit_cyclomort function estimates the parameters of the aforementioned Cauchy distribution using maximum likelihood estimation.

```{r}
T.morts1 <- simPeriodicMorts(300, period = 1, 
                             peaks = c(0.25, 0.75), 
                             rhos = c(0.8, 0.5), 
                             weights = 0.3, dt = .01, 
                             A = .02, plotme = FALSE)

p0 = c(A = 0.05, peak1 = 0.25, peak2 = 0.75, rho1 = 0.5, rho2 = 0.5, weight1 = 0.5)
##MLE for parameters based on simulated data
fits = fit_cyclomort(T.morts1, p0, dt = .01, period = 1)
getCIs(fits)
##Actual parameter values from simulated data
attributes(T.morts1)
```

Here we observe that the estimates for the peaks and weights are quite accurate, implying that the fit_cyclomort function can effectively estimate the peaks and windows of mortality over the course of a year.

##The simPeriodicMorts function: verificiation of fit accuracy

Using the simPeriodicMorts function, we can verify that the fit_cyclomort function provides accurate estimates for each parameter. This function allows us to produce a set of random "Surv" objects that follow the distribution described by the set of input parameters. By producing a set of random mortalities we can observe what a mortality function with the specified distribution might actually look like.

```{r, fig.width = 7, fig.height = 5, fig.align = "center"}

T.morts1 <- simPeriodicMorts(300, period = 1, 
                             peaks = c(0.25, 0.75), 
                             rhos = c(0.8, 0.5), 
                             weights = 0.3, dt = .01, 
                             A = .02, plotme = TRUE)

```
